Designing Qusecs Optimally: Quasi-uniform or Self-similar Layered Materials as Recursively Decomposed Solutions of Geometric Constraint Systems

### 2015-05-16 doc/GMP2015/revision_2/s04_recombination.tex
Revision 2. 24 hours to submission.

The main barrier in recombination when given an optimal DR-plan (of smallest possible size or max fan-in),  is that the number of children (resp. number of roots)---and correspondingly the  size and complexity of the (indecomposable) algebraic system $I$ to be solved---could be arbitrarily large as a function of the size of the input system. This difficulty can persist even after optimal parametrization of the indecomposable system $I$ as in \cite{sitharam2010optimized} to minimize its algebraic complexity.

In the following, we formulate the problem of \dfn{optimal modification} of the indecomposable algebraic system $I$ into a decomposable system with a small DR-plan (low algebraic complexity). Leveraging recent results on \dfn{Cayley configuration spaces}, our approach to the optimal modification problem achieves the following:

\begin{enumerate}[(a)]
    \item
    \header{Small DR-plan.}
    We obtain a  parameterized family of systems $I_{\lambda_F}$---one for each value $\lambda_F$ for the parameters $F$,  all of which have small DR-plans. Thus, given a value $v$ for $\lambda_F$, the system $I_v$ can potentially be solved or realized easily once the orientation type of the solution is known  (when the DR-plan size is small enough).

    \item
    \header{Solution preservation.}
    Moreover, the union of solution spaces of the systems in the family $I_{\lambda_F}$ is guaranteed to contain all of $I$'s solutions.

    \item
    \header{Efficient search.}
    Finally, this union can be projected into the so-called \dfn{Cayley} or distance parameter space $\lambda_F$ that is convex or otherwise easy to traverse in order to search for $I$'s solution (or connected component) of the desired orientation type. For the case when the modification is bounded, this approach provides an efficient algorithm for recombination. We first define the decision version of the problem of optimal modification for decomposition. The standard optimization versions are straightforward.
\end{enumerate}

\problemstatement{Optimal Modification for Decomposition (OMD) Problem.}
Given a generically independent graph $G = (V,E)$ with no non-trivial proper isostatic subgraph (indecomposable) and 2 constants $k$ and $s$, does there exist a set of at most $k$  edges $E_1$ and a set of non-edges $F$ such that the modified graph $H = (V, E\setminus E_1 \cup F)$ has a DR-plan of size at most $s$?  The OMD$_k$ problem is  OMD where $k$ is a fixed bound (not part of the input).   We say that such a tuple $(G,s)$ is a member of the set OMD$_k$. We loosely refer to graphs $G$ as OMD with appropriately small $k,s$ or OMD$_k$ graphs with appropriately small $s$.

It is immediately clear that indecomposable graphs $G$ that belong in OMD$_k$ for small $k$ and $s$  lend themselves to modification  into decomposable graphs satisfying Criteria (a) and (b) above. However, it is not clear how Criterion (c) is met by OMD graphs.

\subsection{Previous Work}
% \todo{We now briefly survey existing techniques for studying 2D qusecs,  many of which are \dfn{bar-joint} systems (Examples 1 and 2 above, see Sections \ref{sec:prelim}, \ref{sec:DRP}, and \ref{sec:recomb}), \dfn{body-hyperpin} systems (Example 4 and 5, see Section \ref{sec:bodypin}) or \dfn{pinned-line incidence} systems (Example 3, see Section \ref{sec:pinnedline}). The limitations of these techniques directly motivate the contributions of this paper.}
We now briefly survey existing techniques for handling the complexity of recombination of 2D DR-plans. The limitations of these techniques directly motivate the contributions in this section.

\begin{enumerate}
    \item
    \header{Optimal Recombination and Solution Space Navigation.}
    For the entire DR-plan, finding all desired solutions is barely tractable even if recombination of solved subsystems is easy for each indecomposable parent system in the DR-plan. This is because even for the simplest, highly decomposable systems with small DR-plans, the problem of finding even a single solution to the input system at the root of the DR-plan is NP-hard  \cite{saxe1979embeddability} and there is a combinatorial explosion of solutions \cite{borcea2004number}. Typically, however, the desired solution has a given orientation type, in which case, the crux of the difficulty is concentrated in the algebraic complexity of (re)combining child system solutions to give a solution to the parent system at any given level of the DR-plan. For fairly general 3D constraint systems, there are algorithms with formal guarantees that uncover underlying matroids to combinatorially obtain an optimal parameterization to minimize the algebraic complexity of the indecomposable parent (sub)systems that occur in the DR-plan \cite{sitharam2010optimized,sitharam2006well,sitharam2010reconciling}, provided the DR-plan meets some of the abovementioned criteria.

    However, the generality of these algorithms trades-off against efficiency, and despite the optimization, the best algorithms can still take exponential time in the number of child subsystems (which can be arbitrarily large even for optimal DR-plans) in order to guarantee all solutions of a given orientation type, even for a single (sub)system in a DR-plan. They are prohibitively slow in practice. We note that, utilizing the DR-plan and optimal recombination as a principled basis, high performance heuristics and software exists \cite{sitharam2006solution} to tame combinatorial explosion via user intervention.

    \item
    \header{Configuration Spaces of Underconstrained Systems.}
    For underconstrained 2D bar-joint and body-hyperpin qusecs obtained from various subclasses of tree-decomposable systems, algorithms have been developed to complete them into isostatic systems \cite{joan-arinyo2003transforming,sitharam2005combinatorial,gao2006ctree,sitharam2010convex} and to find paths within the connected components \cite{sitharam2011cayleyI,hidalgo2011reachability} of standard Cartesian configuration spaces. Most of the algorithms with formal guarantees leverage Cayley configuration space theory \cite{sitharam2010convex,sitharam2011cayleyI,sitharam2011cayleyII} to characterize subclasses of graphs and additional constraints that control combinatorial explosion, and provide faithful bijective representation of connected components and paths. These algorithms have decreasing efficiency as the subclass of systems gets bigger, with highest efficiency for underlying partial 2-tree graphs (alternately called tree-width 2, series-parallel, and $K_4$ minor avoiding), moderate efficiency for 1 degree-of-freedom (dof) graphs with low Cayley complexity (which include common linkages such as the Strandbeest, Limacon and Cardioid), and decreased efficiency for general 1-dof tree-decomposable graphs. While software suites exist  \cite{keycurriculum1995geometer,porta2014open,siemens1999d,todd2007geometry}, no such formal algorithms and guarantees are known for non-tree-decomposable systems.
\end{enumerate}

\subsection{Using Convex Cayley Configuration Spaces}
\label{sec:2-tree-reduction}
%
Next we provide the necessary background to describe a specific approach for achieving the requirements (a)--(c) mentioned above, by restricting the class of reduced graphs $G' = G\setminus E_1$ and their isostatic completions $H$ in the above definition of the OMD problem, and using a key theorem of Convex Cayley configuration spaces \cite{sitharam2010convex}. This theorem characterizes the class of graphs $H$ and non-edges $F$ (Cayley parameters), such that the set of vectors $\lambda_F$ of  attainable lengths of the non-edges $F$ is always convex for any given lengths $\delta$ for the edges of $H$ (i.e, over all the realizations of the bar-joint constraint system or linkage $(H,\delta)$ in 2 dimensions). This set is called the (2-dimensional) \dfn{Cayley configuration space} of the linkage $(H,\delta)$ on the Cayley parameters $F$, denoted $\Phi_F(H,\delta)$ and can be viewed as a ``projection'' of the cartesian realization space of $(H,\delta)$ on the Cayley parameters $F$. Such graphs $H$ are said to have \dfn{convexifiable Cayley configuration spaces with parameters $F$} (short: \dfn{convexifiable}).

To state the theorem, we first have to define the notion of \dfn{2-sums} and \dfn{2-trees}. Let $H_1$ and $H_2$ be two graphs on disjoint sets of vertices $V_1$ and $V_2$, with edge sets $E_1$ and $E_2$ containing edges $(u,v)$ and $(w,x)$ respectively.
% Let $H_1=(V_1,E_1)$ and $H_2=(V_2,E_2)$ be two graphs on disjoint sets of vertices that contain the edges $(u,v)$ and $(w,x)$ respectively.
A \dfn{2-sum} of $H_1$ and $H_2$ is a graph $H$ obtained by taking the union of $H_1$ and $H_2$ and identifying $u=w$ and $v=w$. In this case, $H_1$ and $H_2$ are called \dfn{2-sum components} of $H$. A \dfn{minimal 2-sum component} of $H$ is  one that cannot be further split into 2-sum components. A \dfn{2-tree} is recursively obtained by taking a 2-sum of 2-trees, with the base case of a 2-tree being a triangle. A \dfn{partial 2-tree} is a 2-tree minus some edges. Partial 2-trees have an alternate characterization as the graphs that avoid $K_4$ minors, and are also called series parallel graphs.

@@

\begin{proof}
    The proof follows from the fact that 2-trees are well decomposable and have simple DR-plans of size 2. We know that $G$ can be reduced by removing $k$ edges to create a partial-2-tree $G'$ which can then be completed to an (isostatic) 2-tree by adding some set of non-edges $F$. Thus the modified graph $H = G'\cup F$ has  a DR-plan of size 2, proving the theorem.
\end{proof}

@@

\sidenote{A major advantage of the convex Cayley method is that it is completely unaffected when $\delta$ are intervals rather than exact values~~\cite{sitharam2010convex}.}
%cite my paper with Heping!!!

\myexample
[Using Cayley configuration space].
A graph $G=K_{3,3}$  cannot be decomposed into any nontrivial isostatic graphs, i.e., its DR-plan has a root and 9 children corresponding to the 9 edges. Solving or recombining the system $(G,\delta)$ corresponding to the root of this DR-plan involves solving a quadratic system with 8 equations and variables. Instead of simultaneously solving this system, we could instead use the fact that $G=K_{3,3}$ is in OMD$_2$: remove the edges $e_1,e_2$ in Figure \ref{fig:omd_k33_example} to give a partial 2-tree $G'$. Now add the non-edges $f_1,f_2$ to give a 2-tree $H$ with a DR-plan of size 2. The Cayley configuration space $\Phi_f(G', \delta_{E\setminus e})$ is a single interval of attainable length values $\lambda_F$ for the edge $f$. When $\delta$ is generic, i.e, does not admit collinearities or coincidences in the realizations of $(G,\delta)$, the realization space of $(H, \left<\delta_{E\setminus e}, \lambda_f\right>)$ has 16 solutions $q^p_{\lambda_f}$ (modulo orientation preserving isometries), with distinct orientation types $\sigma_p$ (two orientation choices for each of the 4 triangles) that can be obtained by solving a sequence of 4 single quadratics in 1 variable (DR-plan of size 2). By subdivided binary search in the interval $\lambda_f \in \Phi_f(G', \delta_{E\setminus e})$, the desired solution $p$  of $(G,\delta)$ is found when the length of the nonedge $e$ in  the realization $q^p_{\lambda_f}$ is $\delta_e$.

@@

\subsection{Optimized Modification by Enlarging the class of Reduced
Graphs}
\label{sec:tdecomp}
%
It is possible in principle to decrease $k$ for a OMD$_k$ graph (i.e, the number of edges to be removed to ensure an isostatic completion that is decomposable with a small DR-plan) by considering reduced graphs $G'$ (and modified graphs $H$) that come from a larger class than partial 2-trees but nevertheless have convex Cayley configuration spaces at least when the realization space is restricted to a sufficiently comprehensive orientation type. In particular, the so-called \dfn{tree-decomposable graphs of low Cayley complexity} \cite{sitharam2011cayleyI,sitharam2011cayleyII} include the partial 2-trees and many others that are not partial 2-trees. See an illustration in Figure \ref{fig:omd_k33_example}. These too result in DR-plans of size 2 or 3, putting $G$ in the class OMD$_k$ and thus meeting Criteria (a) and (b). The Criterion (c) is met---for example when $k=1$---because 1-dof Cayley configuration spaces of linkages based on such graphs $G'$ are known to be single intervals when a comprehensive orientation type $\sigma_p$ of the sought solution $p$ is given. In addition, the bounds of these intervals are of low algebraic complexity. More precisely, the bounds  can themselves be computed using a DR-plan of size 2 or 3, i.e., the computation of these bounds is tree-decomposable. Alternatively, the bounds are in a simple quadratic or radically solvable extension field over the rationals, or can be computed by solving a triangularized system of quadratics.

>>>>

The main barrier in recombination when given an optimal DR-plan  (of smallest possible size or max fan-in) for a system $S$,  is that the number of children of the root  (resp. number of roots of the DR-plan)---and correspondingly the  size and complexity of the (indecomposable) algebraic system $I$ to be solved---could be arbitrarily large as a function of the size of $S$. This difficulty can persist even after optimal parametrization of the indecomposable system $I$ as in \cite{sitharam2010optimized} to minimize its algebraic complexity.

\subsection{Previous Work}
% \todo{We now briefly survey existing techniques for studying 2D qusecs,  many of which are \dfn{bar-joint} systems (Examples 1 and 2 above, see Sections \ref{sec:prelim}, \ref{sec:DRP}, and \ref{sec:recomb}), \dfn{body-hyperpin} systems (Example 4 and 5, see Section \ref{sec:bodypin}) or \dfn{pinned-line incidence} systems (Example 3, see Section \ref{sec:pinnedline}). The limitations of these techniques directly motivate the contributions of this paper.}
We now briefly survey existing techniques for handling the complexity of recombination of DR-plans for qusecs. The limitations of these techniques directly motivate the contributions in this section.

\subsubsection{Optimal Recombination and Solution Space Navigation.}
For the entire DR-plan, finding all desired solutions is barely tractable even if recombination of solved subsystems is easy for each indecomposable parent system in the DR-plan. This is because even for the simplest, highly decomposable systems with small DR-plans, the problem of finding even a single solution to the input system at the root of the DR-plan is NP-hard  \cite{saxe1979embeddability} and there is a combinatorial explosion of solutions \cite{borcea2004number}. Typically, however, the desired solution has a given orientation type, in which case, the crux of the difficulty is concentrated in the algebraic complexity of (re)combining child system solutions to give a solution to the parent system at any given level of the DR-plan. For fairly general 3D constraint systems, there are algorithms with formal guarantees that uncover underlying matroids to combinatorially obtain an optimal parameterization to minimize the algebraic complexity of the indecomposable parent (sub)systems that occur in the DR-plan \cite{sitharam2010optimized,sitharam2006well,sitharam2010reconciling}, provided the DR-plan meets some of the abovementioned criteria.

However, the generality of these algorithms trades-off against efficiency, and, despite the optimization, the best algorithms can still take exponential time in the number of child subsystems (which can be arbitrarily large even for optimal DR-plans) in order to guarantee all solutions of a given orientation type, even for a single (sub)system in a DR-plan. They are prohibitively slow in practice. We note that, utilizing the DR-plan and optimal recombination as a principled basis, high performance heuristics and software exists \cite{sitharam2006solution} to tame combinatorial explosion via user intervention.

\subsubsection{Configuration Spaces of Underconstrained Systems.}
For underconstrained 2D bar-joint and body-hyperpin qusecs obtained from various subclasses of tree-decomposable systems, algorithms have been developed to complete them into isostatic systems
\cite{joan-arinyo2003transforming,sitharam2005combinatorial,gao2006ctree,sitharam2010convex} and to find paths within the connected components \cite{sitharam2011cayleyI,hidalgo2011reachability} of standard Cartesian configuration spaces. Most of the algorithms with formal guarantees leverage Cayley configuration space theory \cite{sitharam2010convex,sitharam2011cayleyI,sitharam2011cayleyII} to characterize subclasses of graphs and additional constraints that control combinatorial explosion, and provide faithful bijective representation of connected components and paths. These algorithms have decreasing efficiency as the subclass of systems gets bigger, with highest efficiency for underlying partial 2-tree graphs (alternately called tree-width 2, series-parallel, and $K_4$ minor avoiding), moderate efficiency for 1 degree-of-freedom (dof) graphs with low Cayley complexity (which include common linkages such as the Strandbeest, Limacon, and Cardioid), and decreased efficiency for general 1-dof tree-decomposable graphs. While software suites exist  \cite{keycurriculum1995geometer,porta2014open,siemens1999d,todd2007geometry}, no such formal algorithms and guarantees are known for non-tree-decomposable systems.

\subsection{Optimal Modification for Recombination}
In the following, we formulate the problem of \dfn{optimal modification} of an indecomposable algebraic system $I$ at some node of a (possibly optimal) DR-plan into a decomposable system with a small DR-plan (low algebraic complexity). Leveraging recent results on \dfn{Cayley configuration spaces}, our approach to the optimal modification problem achieves the following:

\begin{enumerate}[(a)]
    \item
    {\sl Small DR-plan.}
    We obtain a  parameterized family of systems $I_{\lambda_F}$---one for each value $\lambda_F$ for the parameters $F$,  all of which have small DR-plans. Thus, given a value $v$ for $\lambda_F$, the system $I_v$ can potentially be solved or realized easily once the orientation type of the solution is known  (when the DR-plan size is small enough).

    \item
    {\sl Solution preservation.}
    Moreover, the union of solution spaces of the systems in the family $I_{\lambda_F}$ is guaranteed to contain all of $I$'s solutions.

    \item
    {\sl Efficient search.}
    Finally,  the so-called \dfn{Cayley} or distance parameter space $\lambda_F$  is convex or otherwise easy to traverse in order to search for $I$'s solution (or connected component) of the desired orientation type. For the case when the modification (number of Cayley parameters) is bounded, this approach provides an efficient algorithm for recombination. We first define the decision version of the problem of optimal modification for decomposition. The standard optimization versions are straightforward.
\end{enumerate}

\problemstatement{Optimal Modification for Decomposition (OMD) Problem.}
Given a generically independent graph $G = (V,E)$ with no non-trivial proper isostatic subgraph (indecomposable) and 2 constants $k$ and $s$, does there exist a set of at most $k$  edges $E_1$ and a set of non-edges $F$ such that the modified graph $H = (V, E\setminus E_1 \cup F)$ has a DR-plan of size at most $s$?  The OMD$_k$ problem is  OMD where $k$ is a fixed bound (not part of the input). We say that such a tuple $(G,s)$ is a member of the set OMD$_k$. We loosely refer to graphs $G$ as OMD with appropriately small $k$ and $s$ or OMD$_k$ with appropriately small $s$.

It is immediately clear that indecomposable graphs $G$ that belong in OMD$_k$ for small $k$ and $s$  lend themselves to modification  into decomposable graphs satisfying Criteria (a) and (b) above. However, it is not clear how Criterion (c) is met by OMD graphs. Before we consider this question, we discuss previous work on recombination of DR-plans.

\subsection{Using Convex Cayley Configuration Spaces}
\label{sec:2-tree-reduction}
%
Next we provide the necessary background to describe a specific approach for achieving the requirements (a)--(c) mentioned above, by restricting the class of reduced graphs $G' = G\setminus E_1$ and their isostatic completions $H$ in the above definition of the OMD problem, and using a key theorem of Convex Cayley configuration spaces \cite{sitharam2010convex}. This theorem characterizes the class of graphs $H$ and non-edges $F$ (Cayley parameters), such that the set of vectors $\lambda_F$ of  attainable lengths of the non-edges $F$ is always convex for any given lengths $\delta$ for the edges of $H$ (i.e.\ over all the realizations of the bar-joint constraint system or linkage $(H,\delta)$ in 2 dimensions). This set is called the (2-dimensional) \dfn{Cayley configuration space} of the linkage $(H,\delta)$ on the Cayley parameters $F$, denoted $\Phi_F(H,\delta)$ and can be viewed as a ``projection'' of the cartesian realization space of $(H,\delta)$ on the Cayley parameters $F$. Such graphs $H$ are said to have \dfn{convexifiable Cayley configuration spaces for some parameters $F$} (short: $H$ is \dfn{convexifiable} \todo{This shortening doesn't exactly make sense}).

To state the theorem, we first have to define the notion of \dfn{2-sums} and \dfn{2-trees}. Let $H_1$ and $H_2$ be two graphs on disjoint sets of vertices $V_1$ and $V_2$, with edge sets $E_1$ and $E_2$ containing edges $(u,v)$ and $(w,x)$ respectively.
% Let $H_1=(V_1,E_1)$ and $H_2=(V_2,E_2)$ be two graphs on disjoint sets of vertices that contain the edges $(u,v)$ and $(w,x)$ respectively.
A \dfn{2-sum} of $H_1$ and $H_2$ is a graph $H$ obtained by taking the union of $H_1$ and $H_2$ and identifying $u=w$ and $v=w$. In this case, $H_1$ and $H_2$ are called \dfn{2-sum components} of $H$. A \dfn{minimal 2-sum component} of $H$ is  one that cannot be further split into 2-sum components. A \dfn{2-tree} is recursively obtained by taking a 2-sum of 2-trees, with the base case of a 2-tree being a triangle. A \dfn{partial 2-tree} is a 2-tree minus some edges. Partial 2-trees have an alternate characterization as the graphs that avoid $K_4$ minors, and are also called series-parallel graphs.

@@

\begin{proof}
    The proof follows from the fact that 2-trees are well decomposable and have simple DR-plans of size 2. We know that $G$ can be reduced by removing $k$ edges to create a partial 2-tree $G'$ which can then be completed to an (isostatic) 2-tree by adding some set of non-edges $F$. Thus the modified graph $H = G'\cup F$ has  a DR-plan of size 2, proving the theorem.
\end{proof}

@@

\sidenote{A major advantage of the convex Cayley method is that it is completely unaffected when $\delta$ are intervals of values rather than exact values~~\cite{sitharam2010convex}.}
%cite my paper with Heping!!!

\myexample
[Using Cayley configuration space].
A graph $G=K_{3,3}$  cannot be decomposed into any nontrivial isostatic graphs, i.e.\ its DR-plan has a root and 9 children corresponding to the 9 edges. Solving or recombining the system $(G,\delta)$ corresponding to the root of this DR-plan involves solving a quadratic system with 8 equations and variables. Instead of simultaneously solving this system, we could instead use the fact that $G=K_{3,3}$ is in OMD$_2$: remove the edges $e_1,e_2$ in Figure \ref{fig:omd_k33_example} to give a partial 2-tree $G'$. Now add the non-edges $f_1,f_2$ to give a 2-tree $H$ with a DR-plan of size 2. The Cayley configuration space $\Phi_f(G', \delta_{E\setminus e})$ is a single interval of attainable length values $\lambda_F$ for the edge $f$. When $\delta$ is generic, i.e.\ does not admit collinearities or coincidences in the realizations of $(G,\delta)$, the realization space of $(H, \left<\delta_{E\setminus e}, \lambda_f\right>)$ has 16 solutions $q^p_{\lambda_f}$ (modulo orientation preserving isometries), with distinct orientation types $\sigma_p$ (two orientation choices for each of the 4 triangles) that can be obtained by solving a sequence of 4 single quadratics in 1 variable (DR-plan of size 2). By subdivided binary search in the interval $\lambda_f \in \Phi_f(G', \delta_{E\setminus e})$, the desired solution $p$  of $(G,\delta)$ is found when the length of the nonedge $e$ in  the realization $q^p_{\lambda_f}$ is $\delta_e$.

@@

\subsection{Optimized Modification by Enlarging the class of Reduced
Graphs}
\label{sec:tdecomp}
%
It is possible in principle to decrease $k$ for a OMD$_k$ graph (i.e.\ the number of edges to be removed to ensure an isostatic completion that is decomposable with a small DR-plan) by considering reduced graphs $G'$ (and modified graphs $H$) that come from a larger class than partial 2-trees but nevertheless have convex Cayley configuration spaces at least when the realization space is restricted to a sufficiently comprehensive orientation type. In particular, the so-called \dfn{tree-decomposable graphs of low Cayley complexity} \cite{sitharam2011cayleyI,sitharam2011cayleyII} include the partial 2-trees and many others that are not partial 2-trees. See an illustration in Figure \ref{fig:omd_k33_example}. These too result in DR-plans of size 2 or 3, putting $G$ in the class OMD$_k$ and thus meeting Criteria (a) and (b). The Criterion (c) is met---for example when $k=1$---because 1-dof Cayley configuration spaces of linkages based on such graphs $G'$ are known to be single intervals when a comprehensive orientation type $\sigma_p$ of the sought solution $p$ is given. In addition, the bounds of these intervals are of low algebraic complexity. More precisely, the bounds  can themselves be computed using a DR-plan of size 2 or 3, i.e.\ the computation of these bounds is tree-decomposable. Alternatively, the bounds are in a simple quadratic or radically solvable extension field over the rationals, or they can be computed by solving a triangularized system of quadratics.

### 2015-05-16 doc/GMP2015/revision_2/s05_problems.tex
Revision 2. 24 hours to submission.

\subsubsection{Special classes of small DR-plans}
As seen in the previous section, 2-trees and tree-decomposable graphs have not only small, but also special DR-plans that permit easy solving---essentially a single quadratic at a time.

@@

We discuss the complexity of the restricted OMD problem (when $k$ is part of the input) in Section \ref{sec:futurework}.

%In the recombination setting, the corresponding \dfn{restricted
%OMD$_k$  problem} requires the reduced graph $G'$ and its isostatic
%completion $H$ to be 2-trees as in Section \ref{sec:2-tree-reduction}
%or to be a low Cayley complexity tree-decomposable graph as in Section
%\ref{sec:tdecomp}. Clearly these problems have deterministic
%polynomial time algorithms, the algorithms run in time exponential in
%$k$. As a result, when $k$ is part of the input, the \dfn{restricted
%OMD problem} has the potential to be difficult. For example, when the
%isostatic completion $H$ is required to be a 2-tree the restricted OMD
%problem is reducible to the maximum spanning series-parallel subgraph
%problem shown by \cite{cai1993spanning} to be NP-complete even if the input
%graph is planar of maximum degree at most 6. However, since the OMD
%problem has other input restrictions such as not having any proper
%isostatic subgraphs, it is not clear if the reverse reduction exists
%and hence it is unclear whether the OMD problem is NP-complete. The
%same holds for the restricted OMD problem where the isostatic
%completion $H$ is required to be a tree-decomposable graph of low
%Cayley complexity (i.e, have special, small DR-plans). One potential
%obstacle to an indecomposable graph $G$'s membership in the restricted
%OMD$_k$ for small $k$ is if $G$ is tri-connected and has large girth.
%In fact, 6-connected (hence rigid) graphs with arbitrarily large girth
%have been constructed in \cite{servatius2000rigidity}.
%
\subsubsection{Optimal Modification, Completion and Recombination: formal
connections}
%
The OMD problem is closely related to a well-studied problem of completion of an underconstrained system to an isostatic one with a small DR-plan.
\begin{observation}\label{obs:OC_to_OMD}
    The (decision version of) the \textbf{optimal completion problem (OC)} from \cite{sitharam2005combinatorial,joan-arinyo2003transforming,zhang-gao2006well} is OMD$_0$.
\end{observation}
In fact, a \dfn{restricted OC} problem was studied by \cite{joan-arinyo2003transforming} requiring the completion to be tree-decomposable.
%It is not clear if
%the reverse reduction holds, i.e, is OMD reducible to OC?

@@

In order to connect the OR problem to OMD, when the input graph is the isostatic graph at the DR-plan root, we do not consider the case where the two child \dfn{solved subgraphs} (corresponding to already solved subsystems) have a nontrivial intersection (in this case the recombination is trivial). We only consider the case where no two child solved subgraphs (resp. two root subgraphs when the input graph is underconstrained) share more than 1 vertex. We replace such solved subgraphs  by isostatic graphs as follows. If a solved subgraph shares at most one vertex with the remainder of the graph, simply replace it by an edge one of whose endpoints  is the shared vertex. Otherwise, replace it by  a 2-tree graph of the shared vertices. Finally, we add the additional restriction to the OM Problem that when any edge in a solved subgraph is chosen among the $k$ edges to be removed, in fact the entire solved subgraph must be removed  and all of its edges must be counted in $k$.

This reduction is used also for transferring  algorithms for optimal DR-planning, recombination, completion, OMD, and other problems from bar-joint systems to so-called \dfn{body-hyperpin}, defined in Section \ref{sec:bodypin}, by showing that the problems for the latter are reduced to the corresponding problems on bar-joint systems.

>>>>

\subsubsection{Special classes of small DR-plans}
As seen in the previous section, 2-trees and tree-decomposable graphs have not only small, but also special DR-plans that permit easy solving---essentially by solving a single quadratic at a time.

@@

We discuss the complexity of the restricted OMD problem (when $k$ is part of the input) in the open-problem Section \ref{sec:futurework}.

%In the recombination setting, the corresponding \dfn{restricted
%OMD$_k$  problem} requires the reduced graph $G'$ and its isostatic
%completion $H$ to be 2-trees as in Section \ref{sec:2-tree-reduction}
%or to be a low Cayley complexity tree-decomposable graph as in Section
%\ref{sec:tdecomp}. Clearly these problems have deterministic
%polynomial time algorithms, the algorithms run in time exponential in
%$k$. As a result, when $k$ is part of the input, the \dfn{restricted
%OMD problem} has the potential to be difficult. For example, when the
%isostatic completion $H$ is required to be a 2-tree the restricted OMD
%problem is reducible to the maximum spanning series-parallel subgraph
%problem shown by \cite{cai1993spanning} to be NP-complete even if the input
%graph is planar of maximum degree at most 6. However, since the OMD
%problem has other input restrictions such as not having any proper
%isostatic subgraphs, it is not clear if the reverse reduction exists
%and hence it is unclear whether the OMD problem is NP-complete. The
%same holds for the restricted OMD problem where the isostatic
%completion $H$ is required to be a tree-decomposable graph of low
%Cayley complexity (i.e, have special, small DR-plans). One potential
%obstacle to an indecomposable graph $G$'s membership in the restricted
%OMD$_k$ for small $k$ is if $G$ is tri-connected and has large girth.
%In fact, 6-connected (hence rigid) graphs with arbitrarily large girth
%have been constructed in \cite{servatius2000rigidity}.
%
\subsubsection{Optimal Modification, Completion and Recombination: previous work and formal
connections}
%
The OMD problem is closely related to a well-studied problem of completion of an underconstrained system to an isostatic one with a small DR-plan.
\begin{observation}\label{obs:OC_to_OMD}
    The (decision version of) the \textbf{optimal completion problem (OC)} from \cite{sitharam2005combinatorial,joan-arinyo2003transforming,zhang-gao2006well} is OMD$_0$.
\end{observation}
In fact, a \dfn{restricted OC} problem was studied by \cite{joan-arinyo2003transforming} requiring the completion to be tree-decomposable.
%It is not clear if
%the reverse reduction holds, i.e, is OMD reducible to OC?

@@

In order to connect the OR problem to OMD, when the input graph is the isostatic graph at the DR-plan root, we do not consider the case where the two child \dfn{solved subgraphs} (corresponding to already solved subsystems) have a nontrivial intersection (in this case the recombination is trivial). We only consider the case where no two child solved subgraphs (resp. two root subgraphs when the input graph is underconstrained) share more than 1 vertex. We replace such solved subgraphs  by isostatic graphs as follows. If a solved subgraph shares at most one vertex with the remainder of the graph, simply replace it by an edge one of whose endpoints  is the shared vertex. Otherwise, replace it by  a 2-tree graph of the shared vertices. Finally, we add the additional restriction to the OM problem that when any edge in a solved subgraph is chosen among the $k$ edges to be removed, in fact the entire solved subgraph must be removed  and all of its edges must be counted in $k$.

This reduction is used also for adapting  algorithms for optimal DR-planning, recombination, completion, OMD, and other problems from bar-joint systems to so-called \dfn{body-hyperpin}, defined in Section \ref{sec:bodypin}, by showing that the problems for the latter are reduced to the corresponding problems on bar-joint systems.
